Review Paper Draft

### 2020-06-29 main.tex
2.3 edits

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. It should be noted that by attempting to reduce the effects of population structure, these methods reduce the effects of environmental confounding by blocking the casual pathway between population structure and observed SNPs.

The genomic control method is a univariate testing procedure which modifies the Cochran-Armitage test for trend by a correction factor \pb{genomic control is not specific to trend tests; it can be used for any type of test statistic inflation}. This correction attempts to quantify and remove the effect of population structure on the distribution of the chi-square statistic using markers unrelated to the phenotype of interest and independent of the marker being tested for association \cite{devlin1999genomic, bacanu2000power, wang2009testing}. The genomic control method retains the disadvantages of univariate SNP testing and its power to detect true associations is likely to suffer in the presence of linkage disequilibrium \pb{reference?}.  \pb{This paragraph is a little too negative about genomic control...needs to be softened slightly}

Stratified analysis attempts to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \cite{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall somewhere between these two extremes \pb{seems useful to bring in idea of a gradient, or continuously varying relatedness, here}. By clustering those individuals within in one subpopulation or another we lose information about more high dimensional relatedness \pb{I think ``high dimensional relatedness'' is confusing/ambiguous here.  What do you mean by it?}.

@@

Like PC-adjustment methods, LMMs rely on a pair-wise similarity matrix \pb{as we discussed, this is a bit odd to say here} to estimate and subsequently correct for relatedness among subjects. However, the term LMM has been used to describe a number of distinct procedures for analyzing structured genetic data which differ in their objectives and methods. There are two main objectives of LMMs used in a structured genetic context: estimating narrow-sense heritability, and estimating individual SNP effects.

>>>>

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. It should be noted that by attempting to reduce the effects of population structure, these methods reduce the effects of environmental confounding by blocking the casual pathway between population structure and observed SNPs. This approach thus seeks to remove the effect of environmental heterogeneity on observed SNPs indirectly.

\anna{DAG here to illustrate this?}

The genomic control (GC) method is a univariate testing procedure which modifies a test statistic by a deflation factor. This deflation factor attempts to quantify and remove the effect of population structure on the distribution of the appropriate test statistic, and is estimated using loci independent of the phenotype of interest and the loci being tested for association \cite{devlin1999genomic, bacanu2000power, wang2009testing}. However, this deflation factor has been reported to be sensitive to the nature and quantity of the loci used to estimate it \cite{hellwege2017population, marchini2004effects}. Additionally, GC does not correct effect size estimates, so resulting coefficient estimates will be unreliable after GC is applied, even if their corresponding test statistics and p-values have been appropriately adjusted \cite{hellwege2017population}.

Stratified analysis attempts to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \cite{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall between these two extremes and are characterized by continuously varying relatedness and a gradient of air quality. By clustering those individuals within in one subpopulation or another we lose information about their fine-grained, subtle relatedness.

@@

% rely on a pair-wise similarity matrix
Like PC-adjustment methods, LMMs utilize the left singular vectors of the design matrix to estimate and subsequently correct for relatedness among subjects. However, the term LMM has been used to describe a number of distinct procedures for analyzing structured genetic data which differ in their objectives and methods. There are two main objectives of LMMs used in a structured genetic context: estimating narrow-sense heritability, and estimating individual SNP effects.
